Random Matrix Theory Tools for the Analysis of Functional Magnetic-Resonance Imaging Examinations

### 2022-10-06 paper/bookdown/rmt.Rmd
update

\input{preamble.tex}

\title{Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations}
\author{Derek M. Berger \and Gurpreet S. Matharoo \and Jacob Levman}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
    Previous studies have investigated the potential of using analytic techniques from Random Matrix
    Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
    application of RMT-based features for the analysis of functional MRI (fMRI) across diverse
    datasets. As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain
    features extracted from voxel-wise functional connectivity to contain information useful for
    classifying between various psychological processes, (2) assess these features’
    predictive—rather than explanatory—value, (3) investigate the effect of varying RMT analysis
    methods on the robustness of study findings, and (4) make general-purpose code publicly
    available for users to extract these features from a wide variety of data. We find preliminary
    evidence suggesting that RMT-inspired features may have unique potential in analyses of fMRI
    functional connectivity.

    \subparagraph{Index Terms:} feature extraction, functional connectivity, functional magnetic
    resonance imaging, random matrix theory, voxelwise connectivity

    \subparagraph{Authors}
    Derek Berger is with the department of Computer Science, St. Francis Xavier University, Antigonish
    NS B2G 2W5, Canada (e-mail: dberger@stfx.ca).

@@

    Jacob Levman is with the Department of Computer Science, St. Francis Xavier University,
    Antigonish NS B2G 2W5, Canada (e-mail: jlevman@stfx.ca).
\end{abstract}

\section{Introduction}

@@

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous
(\cite{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}).
This complexity suggests fMRI is a candidate to be studied using Random Matrix Theory (RMT), a set
of mathematical tools originally developed some 50 years ago to solve complex problems in nuclear
physics (\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}). In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

\begin{sloppypar}
These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has been applied to atmospheric
physics (\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations \cite{plerouRandomMatrixApproach2002} , social networks \cite{jalanUncoveringRandomnessSuccess2014}, random networks \cite{bandyopadhyayUniversalityComplexNetworks2007},
network-formation in liquids (\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009})  and amorphous clusters (\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}). Within biological systems,
RMT has also been used to successfully model aspects of  amino acid functional relationships (\cite{bhadolaTargetingFunctionalMotifs2016}),
synchrony in epileptic seizures (\cite{osorioPhasesynchronizationRandommatrixBased2011}), and in protein-protein interactions both in different species (\cite{agrawalQuantifyingRandomnessProtein2014}) and breast cancer (\cite{raiRandomnessPreservedPatterns2015}).
\end{sloppypar}

In recent years, RMT analyses have been applied to study brain functioning.  The earliest  study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) (\cite{sebaRandomMatrixAnalysis2003}).
Somewhat more recently, RMT was used to evaluate the quality of whole brain features extracted from
fMRI data (\cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}). RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
(\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}).

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states (\cite{wangSpectralPropertiesTemporal2015}), between
subjects with and without ADHD (\cite{wangRandomMatrixTheory2016}), and between pain and non-pain
states (\cite{matharooSpontaneousBackpainAlters2020}). Across these three studies, the spectra of
resting or low-attention states exhibited properties closest to the GOE.

@@

\section{Materials and Methods}

\subsection{Datasets}

@@

However, since previous studies employing RMT for feature extraction
(\cite{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}) have partially interpreted
their findings with respect to attentional processes, we have elected to use datasets that involve
attention in some way. This includes measures of attention or vigilance, or comparisons between a
condition involving more attention and focus (e.g. a task state) or conditions which should involve
considerably less attention or focus (e.g. resting-state).

\begin{table}[h!]
\small
\centering
\begin{tabular}{ c c c c c c c }
\hline
\textbf{ID}  & \textbf{FOV (mm)}  & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{Time (s)} \\
\hline
PSYCH   &  192              & 128 x 128 x 70 & 1.5 x 1.5 x 1.5 & 3.0 & 300 & 15 \\
OSTEO   &  -                & 64 x 64 x 36   & 3.4 x 3.4 x 3.0 & 2.5 & 300 & 12.5 \\
REFLECT &  210              & 72 x 72 x 46   & 3.0 x 3.0 x 3.0 & 2.5 & 204 & 10.2 \\
REFLECT &  240              & 64 x 64 x 33   & 3.8 x 3.8 x 3.8 & 2.0 & 260 & 8.7 \\
PARK    &  240 x 240 x 129  & 80 x 80 x 43   & 3.0 x 3.0 x 3.0 & 2.4 & 149 & 5.96 \\
LEARN   &  -                & 64 x 64 x 36   & 3.0 x 3.0 x 3.0 & 2.0 & 196 & 6.5 \\
\hline
\end{tabular}
\caption{
    ID = Identifier for paper. FOV = Field of View. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P slices with dimensions M x N.
}
\label{table:1}
\end{table}

Dataset scan parameters and acquisition details are summarized in Table I, and sample and subgroup
sizes are summarized in Table II. For all datasets, we use the raw (unpreprocessed) scans, and
perform our own preprocessing (detailed later). Given the complexity of these datasets, we highlight
below only those aspects relevant to the current paper.

\begin{table}[h!]
\small
\centering
\begin{tabular}{ l l c c c }
\hline
\textbf{ID}  & \textbf{Subgroup / Condition}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
PSYCH   &  high\_vigilance    & 12 / 11 & resting-state                    & 2 \\
        &  low\_vigilance     & 10 / 11 & resting-state                    & 2 \\
        &  trait\_attend      & 12 / 15 & resting-state                    & 2 \\
        &  trait\_non-attend  & 10 / 7  & resting-state                    & 2 \\
        &  task\_attend       & 13 / 15 & resting-state                    & 2 \\
        &  task\_nonattend    &  9 / 7  & resting-state                    & 2 \\
\hline
LEARN   &  task              & 24      & learn image sketches & 16 \\
        &  rest              & 24      & resting-state                    & 2 \\
\hline
REFLECT &  control (no pain) & 20      & resting-state                    & 1 \\
        &  duloxetine        & 19      & resting-state                    & 1 \\
        &  placebo           & 37      & resting-state                    & 1 \\
        &  allpain (duloxetine + placebo) & 56         & resting-state    & 1 \\
\hline
PARK    &  Parkinson's       & 17      & ANT                              & 6 \\
        &  control           & 17      & ANT                              & 6 \\
\hline
ECHO    &  summed/interleave & 31      & resting-state                    & 1 \\
        &  summed/interleave & 31      & imagery/reflection               & 4 \\
\hline
\end{tabular}
\caption{
    ID = Identifier for paper. FOV = Field of View. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P slices with dimensions M x N.
}
\label{table:2}
\end{table}

\subsubsection{Dataset 1}
Dataset 1 (referred to in Fig. and tables with the label ``PSYCH'') has detailed and extensive
information available in \cite{gorgolewskiHighResolution7Tesla2015}. Subjects completed a battery of
psychological measurements, and various resting-state fMRI (rs-fMRI) images were obtained. Only the
whole-brain 7-Tesla rs-fMRI scans were used for the current paper, to make the results comparable to
the other included datasets. For the purpose of keeping with the ``attention'' theme of the analyses
in this paper, three pairings of subgroups were constructed via z-score splits of the various
sum-scores of scale items involving concepts closely related to attention.

\paragraph{Trait Attention}
The PANAS-X (\cite{watsonPANASXManualPositive1994}) is a 9-point Likert-type self-report measure
that queries the extent to which subjects agree particular affective terms have applied to them over
a recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) sum-score was constructed from
the scale items: ``attentive'', ``alert'', ``interested'', ``concentrating'', and reverse-scored items
``sluggish'', ``tired'', ``sleepy'', ``drowsy''. Subjects with a z-score greater / less than zero on this
sum-score were placed in the high / low ``trait attention'' category

\paragraph{Task Attention}
The Conjunctive Continuous Performance Task (CCPT;
\cite{shalevConjunctiveContinuousPerformance2011}) is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low ``task attention'' category.

\paragraph{Vigilance}
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. ``my thoughts involved other people'', ``I thought about something positive'' (translated
from German; \cite{gorgolewskiCorrespondenceIndividualDifferences2014}). One question assesses
alertness (``I was fully awake'' / ``war ich vollkommen wach'') and was used to create high / low
``vigilance'' conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in Fig. and tables with the label ``OSTEO'') is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls (``nopain''
condition) and individuals with knee osteoarthritis (``pain'' condition) treated for 2 weeks with
either placebo (``pain'' condition) or duloxetine (``duloxetine'' condition). An additional grouping was
also created with both the placebo and duloxetine data (``allpain'' condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table I. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

\subsubsection{Dataset 3}
Dataset 3  (referred to here with the label ``REFLECT'', due to the ``reflective” nature of the
cognitive tasks) is described in full detail by \cite{dupreMultiechoFMRIReplication2016}. Both
multi-echo rs- and task-fMRI scans were conducted with scan parameters summarized in Table II. Tasks
involved viewing various affective images (see \cite{sprengPatternsBrainActivity2010} for details)
and performing tasks related to autobiographical memory, prospection, or theory-of-mind reasoning.
Continuing with the theme of broadly comparing high versus low attention throughout this paper, the
different tasks were simply treated generically as ``attending'', in contrast to the non-attentive
resting-state scans. This resulted in four task state scans and one resting-state scan per subject.

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the ``summed'' condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast (\cite{kunduMultiechoFMRIReview2017}). In the ``interleave''
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's ith signal value
from echo j as , then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

This creates a physically-meaningful 4D image that contains the full, original multi-echo fMRI
time-series data without any alterations. It should be noted that the purpose of these combination
methods is not to create a 4D volume that allows optimal detection of the purely task-linked BOLD
signal, but to ensure all scan information was used with minimal additional analytic assumptions.

\subsubsection{Dataset 4}
Dataset 4 (referred to in Fig. and tables with the label ``PARK'' or ``PARKINSONS'') is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table I. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task (ANT)
(\cite{fanActivationAttentionalNetworks2005}). Subgroup details are available in Table I. As the ANT
task demands high attention, it is not as straightforward to create ``high'' or ``low'' attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in Fig. and tables with the label ``LEARN'') is documented and described in
detail in \cite{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components (see Fig. 1 of
\cite{schapiroHumanHippocampalReplay2018}). After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention ``task'' subgroup for the purposes of the
analysis of this paper. Resting state scans were included in the low-attention ``rest'' subgroup.

\subsection{Preprocessing}
Some consideration should be given to whether it is appropriate to apply standard preprocessing
steps before extracting RMT features, as preprocessing steps should be developed, tested, and
employed always with respect to a set of inferential goals and a specific analytic framework.
However, these preprocessing steps are increasingly recognized to have complex effects on fMRI
analyses (\cite{parkerBenefitSliceTiming2019}). For example, spatial smoothing has clear benefits
when employing a general linear model
(\cite{miklEffectsSpatialSmoothing2008,alakorkkoEffectsSpatialSmoothing2017,worsleyAnalysisFMRITimeSeries1995}),
but is less motivated in functional network analyses with small ROIs, where the effect of smoothing
can be considerable
(\cite{alakorkkoEffectsSpatialSmoothing2017,goelmanMaximizingNegativeCorrelations2014,trianaEffectsSpatialSmoothing2020},
with these effects interacting complexly with factors like field strength and image resolution
(\cite{triantafyllouEffectSpatialSmoothing2006}). Similarly, image registration is usually required
to allow comparisons across (or even within) subjects, but the choice of registration strategy can
have dramatic effects on functional network analyses (\cite{magalhaesImpactNormalizationSegmentation2014}).

The RMT features extracted here are based on functional correlations rather than ROI activations,
and are completely insensitive to voxel location (spatial re-arrangement of an image’s voxels does
not change the eigenvalues of the correlation matrix). Thus RMT features summarize all voxel-wise
correlations in a whole-brain and location-free manner, which brings into question some of the usual
motivations for performing spatial smoothing, or registration to identify ROIs. In addition, if
image dimensions are identical between subjects, the RMT features (by virtue of their definition and
the unfolding procedure) already allow for direct comparisons. Thus, the usual motivations for
registration are also no longer clear when extracting RMT features.

Given these particularities, as well as the novelty of the RMT-based methods and the diversity of
the datasets, we elected to minimize the complexity of the preprocessing pipeline to ensure that
data processing is kept as standardized as possible. Ultimately, different preprocessing pipelines
for RMT fMRI analyses can be expected to have different benefits and shortcomings depending on the
question being investigated, and possibly also the specific RMT features being relied upon, but such
an investigation is beyond the scope of the current study.

For this paper’s analyses, brain extraction is necessary to limit analyses to the appropriate
voxels, motion correction is necessary to allow proper brain extraction, and both motion-correction
and slice-time correction simplify the interpretation of voxel-voxel correlations across voxel
locations. Whereas physiological noise regression and bias field correction would perhaps be
appropriate here, there is insufficient data to perform these steps for all datasets (Table II).
Thus, motion-correction was performed with FSL's MCFLIRT
(\cite{jenkinsonImprovedOptimizationRobust2002}), slicetime correction with the default method in
FSL, and brain extraction was performed with BET (\cite{smithFastRobustAutomated2002}). Due to
differences in the size of the REFLECT data, the higher-resolution resting-state data was
interpolated to have the same spatial dimensions as the task data.

\subsection{RMT-Derived Features}
This paper examines features informed by various aspects of RMT, and which can be extracted from
empirical correlation matrices. To keep with previous studies, we limit our analyses to features
extracted only from the eigenvalues of the matrices of correlations between all voxel signals.

Two classes of features were examined in this study: those that require unfolding, and those that do
not. In order to compare the observed eigenvalue distribution to some theoretical predictions, the
eigenvalues must be unfolded (\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}).
The details and motivation behind the unfolding procedure are well documented in
\cite{guhrRandommatrixTheoriesQuantum1998a}. Practically, the unfolding process can be viewed as a
smoothing and rescaling procedure where the originally observed spectrum \(\lambda_1 \le \dots \le
\lambda_n\) is smoothly mapped to the unfolded spectrum \(e_1, \dots, e_n\), such that \(\sum d_i/n
\approx 1\), if \(d_i = e_{i+1} - e_i \). For most empirical data, the true smoothing function is
not known, and so must be approximated
(\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}). Typically, a polynomial is
chosen (\cite{abul-magdUnfoldingSpectrumChaotic2014}).

A number of statistics (or ``spectral observables''; \cite{mehtaRandomMatrices2004}) can then be
computed from this unfolded spectrum. In the literature, the two most common points of comparison
for these unfolded features are typically the features of the Poisson or Gaussian Diagonal Ensemble
(GDE) and the GOE (see e.g.
\cite{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016}). The
GOE (or GDE) consists of real, symmetric matrices where entries (or only diagonal entries) are
normal random variables with zero mean and unit variance. When the empirically observed features are
close to those of the GOE, this is usually interpreted as indicating a high degree of randomness and
statistical independence in the data, and when the observed features are closer to those expected in
the GDE or Poisson case, this is interpreted as indicating a high degree of coherence or dependence
(\cite{santhanamStatisticsAtmosphericCorrelations2001,jalanUncoveringRandomnessSuccess2014,bandyopadhyayUniversalityComplexNetworks2007,agrawalQuantifyingRandomnessProtein2014,raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}).

The Nearest-Neighbours Spacings Distribution (NNSD) is the most basic RMT feature, and is the
distribution of the spacings (differences) of adjacent unfolded eigenvalues. Normally, the NNSD is
analysed visually via a histogram of the distribution of spacings, however it is also possible to
analyse it more formally by fitting the Brody distribution (\cite{brodyStatisticalMeasureRepulsion1973}):

\[
p(s) = \alpha (1 + \beta) s^\alpha \exp({\beta s^{\alpha+1}}),
\quad \alpha = \left[ \Gamma \left( \frac{\beta+2}{\beta+1} \right) \right]^{\beta+1}
\]

where \(\beta\) is free to vary within \([0, 1]\), with \(\beta = 0\)  producing the expected NNSD
for the GDE case, and \(\beta = 1\) producing the expected NNSD for the GOE case. In this paper, we
fit, via maximum likelihood, the empirically observed spacings with the Brody distribution, and
report the Brody parameter, \(\beta\), as an important unfolded feature to investigate.

Whereas the NNSD and Brody parameter characterize the local properties of eigenvalues close on the
spectrum, the long-range properties relating eigenvalues distant on the spectrum can be
characterized by both the spectral rigidity and level number variance
(\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. The spectral rigidity,
\(\Delta_3(L)\), is calculated for any real value \(L\) as:

\[
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\]

where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) \cite{guhrRandommatrixTheoriesQuantum1998a}. The level
number variance, \(\Sigma^2(L)\), is closely related to the spectral rigidity
(\cite{mehtaRandomMatrices2004}), and is calculated as:

\[
\Sigma^2(L) = \left\langle \eta^2(L, c) \right\rangle_c - \left\langle \eta(L, c) \right\rangle^2_c
\]

where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\), and where \(c\), \(L\),
and \(\langle \cdot \rangle_c\) are as above (\cite{guhrRandommatrixTheoriesQuantum1998a}). Both the
spectral rigidity and level number variance are also computed and investigated in this study.

In addition to the unfolded features suggested by RMT, there are some potential ``raw'' features that
do not require an unfolding procedure. The entire eigenvalue spectrum likely contains important
information related to the system, and is also investigated as an RMT-inspired feature.

The distribution of eigenvalues of large rectangular matrices with finite-variance, independent and
identically distributed (\textit{i.i.d.}) entries also converges to the Marchenko-Pastur (MP)
distribution (\cite{marcenkoDISTRIBUTIONEIGENVALUESSETS1967,yaoLargeSampleCovariance2015}):

\[
\rho(\lambda) =
\frac{1}{2\pi\sigma^2}
\frac{\sqrt{(\lambda_+ - \lambda)(\lambda - \lambda_-)}}{r\lambda}, \;
\lambda_{\pm} = \sigma^2\left(1 \pm \sqrt{\lambda}\right)^2
\]

where \(\sigma^2\) is the true variance of the distribution of each i.i.d. entry (equal to 1 in the
case of correlation matrices), and where \(r = N / T\), for matrices of dimension \(N \times T\).
While real-world data will not, in general, be i.i.d., the extent to which spectra overlap with or
are well-described by the MP distribution has been used to guide analytical decisions regarding the
relevance of empirically observed eigenvalues. That is, since the majority of the eigenvalues of
random correlation matrices will fall within \([\lambda_{-}, \lambda_{+}]\), eigenvalues observed to
fall within this region are typically considered to be the result of noise
(\cite{veraartDenoisingDiffusionMRI2016,veraartDiffusionMRINoise2016,ulfarssonDimensionEstimationNoisy2008,macmahonCommunityDetectionCorrelation2015}).
We examine as a potential feature the proportion of eigenvalues that fall within this range, and
call this the noise ratio. In addition, we calculate a broader, shifted noise ratio, based on the
adjustments of \cite{almogUncoveringFunctionalSignature2019} which takes into account the larger
degree of correlation in empirical data.

\subsection{RMT Feature Extraction}
The entire 4-dimensional \(m \times n \times n \times p \times t\) scan array for each run,
excluding voxels with constant signals (which lead to undefined correlations), and excluding voxels
identified as non-brain via preprocessing, was flattened to a two-dimensional \(m \cdot n \cdot p
\times t\) matrix, and then the \(t - 1\) eigenvalues of that correlation matrix were computed. The
value of \(t\) for a scan can be inferred from the number of volumes in Table I. The various RMT
features mentioned above were then extracted via the empyricalRMT library [64] developed and made
publicly available as part of this study, as well as via specific code for this paper (which is
documented and available online (\cite{bergerderekm.DMBergerRandommatrixfmriV02020}).

As both the spectral rigidity and level number variance can be highly sensitive to the choice of
smoothing procedure used in the unfolding method
(\cite{abul-magdUnfoldingSpectrumChaotic2014,moralesImprovedUnfoldingDetrending2011,abueleninSpectralUnfoldingChaotic2018,fossionRandommatrixSpectraTime2013,abueleninEffectUnfoldingSpectral2012}),
all unfolded features in this study were calculated using polynomials of degree 5, 7, 9, 11, and 13
as the smoothing technique. As much of the variance in the unfolded spectral statistics can be due
to the largest eigenvalues acting as anchor points during smoothing
(\cite{moralesImprovedUnfoldingDetrending2011,abueleninEffectUnfoldingSpectral2012}), all unfolded
features were also calculated twice (in addition to each polynomial degree) to examine the effects
of trimming on spectral observables: once with only the largest eigenvalue omitted, and once with
the largest 20 eigenvalues omitted.

>>>>

---
title: "Random Matrix Theory Tools for the Analysis of Functional Magnetic-Resonance Imaging Examinations"
fontsize: 10pt
urlcolor: blue
linkcolor: blue
# mainfont: "Garamond Premier Pro"
# biblio-style: "apalike"
bibliography: "rmt.bib"  # exported from Zotero with better BibLaTeX extension
link-citations: true
output:
    # bookdown::pdf_document2:
    bookdown::pdf_book:
        latex_engine: xelatex
        toc: false
        # toc_depth: 2
        number_sections: false
        highlight: tango
        # https://bookdown.org/yihui/rmarkdown-cookbook/figure-placement.html
        # unicode-math needed for bold italic math
        # (https://tex.stackexchange.com/questions/431013/error-extended-mathchar-used-as-mathchar-when-using-bm)
        extra_dependencies: ["float", "unicode-math"]
        mathjax: default
        includes:
            in_header: preamble.tex
        split_bib: true
        fig_width: 6
        fig_height: 4
        link-citations: true
citation-style: ieee.csl # https://www.zotero.org/styles
---

# Abstract

Previous studies have investigated the potential of using analytic techniques from Random Matrix
Theory (RMT) to investigate magnetic resonance imaging (MRI) data. We assess the potential
application of RMT-based features for the analysis of functional MRI (fMRI) across diverse datasets.
As novel contributions, we (1) assess the potential for RMT-inspired, whole-brain features extracted
from voxel-wise functional connectivity to contain information useful for classifying between
various psychological processes, (2) assess these features’ predictive—rather than
explanatory—value, (3) investigate the effect of varying RMT analysis methods on the robustness of
study findings, and (4) make general-purpose code publicly available for users to extract these
features from a wide variety of data. We find preliminary evidence suggesting that RMT-inspired
features may have unique potential in analyses of fMRI functional connectivity.

## Index Terms / Keywords

feature extraction, functional connectivity, functional magnetic resonance imaging, random matrix
theory, voxelwise connectivity

## Funding Info

Manuscript submitted January 20, 2021. This work was supported by the  Natural Science and
Engineering Research Council of Canada's Canada Research Chair grant (grant number 231266) to JL,
Natural Science and Engineering Research Council of Canada Discovery Grant to JL, a Canada
Foundation for Innovation and Nova Scotia Research and Innovation Trust infrastructure grant
(R0176004) to JL, a St. Francis Xavier University research startup grant to JL (grant number
R0168020), and a St. Francis Xavier University UCR grant to JL.

## Author Info

Derek Berger is with the department of Computer Science, St. Francis Xavier University, Antigonish
NS B2G 2W5, Canada (e-mail: dberger@stfx.ca).

@@

Jacob Levman is with the Department of Computer Science, St. Francis Xavier University, Antigonish
NS B2G 2W5, Canada (e-mail: jlevman@stfx.ca).

\newpage

# Introduction

@@

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous [1]–[4]. This complexity suggests fMRI is a
candidate to be studied using Random Matrix Theory (RMT), a set of mathematical tools originally
developed some 50 years ago to solve complex problems in nuclear physics [5], [6]. In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems [5]–[11], and at larger scales, RMT has been applied to atmospheric
physics [12], stock cross-correlations [13], social networks [14], random networks [15],
network-formation in liquids [16], [17] and amorphous clusters [18]–[21]. Within biological systems,
RMT has also been used to successfully model aspects of  amino acid functional relationships [22],
synchrony in epileptic seizures [23], and in protein-protein interactions both in different species
[24] and breast cancer [25].

In recent years, RMT analyses have been applied to study brain functioning.  The earliest  study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) [26].  Somewhat more recently, RMT was used
to evaluate the quality of whole brain features extracted from fMRI data [27], [28]. RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising [29]–[31].

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states [32], between subjects with and without ADHD [33], and
between pain and non-pain states [34]. Across these three studies, the spectra of resting or
low-attention states exhibited properties closest to the GOE.

@@

# Materials and Methods

## Datasets

@@

However, since previous studies employing RMT for feature extraction [33], [34] have partially
interpreted their findings with respect to attentional processes, we have elected to use datasets
that involve attention in some way. This includes measures of attention or vigilance, or comparisons
between a condition involving more attention and focus (e.g. a task state) or conditions which
should involve considerably less attention or focus (e.g. resting-state).

```{r , echo=FALSE, results='asis'}
  cat(' Table: (\\#tab:scan-params) Summary of Scan Parameters. ID = Identifier for paper. FOV = Field of View. TR = Time of Repetition (seconds). Time = total duration (minutes) of each scan. Dimensions listed as M × N × P, indicate P slices with dimensions M × N.

  |   ID   |     FOV (mm) | Dimensions | Voxel Size (mm) |  TR  | Volumes | Time (m) |
  |:------:|:------------:|:-----------:|:---------------:|:----:|:-------:|:--------:|
  | PSYCH  | 192          | 128×128×70 | 1.5×1.5×1.5     | 3.0  |     300 |  15     |
  | OSTEO  | -            | 64×64×36   | 3.4×3.4×3.0     | 2.5  |     300 |  12.5   |
  | REFLECT| 210          | 72×72×46   | 3.0×3.0×3.0     | 3.0  |     204 |  10.2   |
  |        | 240          | 64×64×33   | 3.8×3.8×3.8     | 2.0  |     260 |  8.7    |
  | PARK   | 240×240×129  | 80×80×43   | 3.0×3.0×3.0     | 2.4  |     149 |  5.96   |
  | LEARN  |  -           | 64×64×36   | 3.0×3.0×3.0     | 2.0  |     195 |  6.5    |')

```

## A Simple Linear MLP

A citation to [@abueleninSpectralUnfoldingChaotic2018a;
@abul-magdModellingHighwaytrafficHeadway2007; @alakorkkoEffectsSpatialSmoothing2017].

As a concrete example, imagine we are build a simple ANN from $\mathbbm{R}$ to $\mathbbm{R}^2$ by
way of two larger intermediate layers. We might have for example

```python
ann = Sequential([
    Linear(1, 2),
    Linear(2, 1),
])
```

which learns the weights

$$
\begin{aligned}
\mathbf{W}_1 &= \begin{bmatrix} w_1 & w_2 \end{bmatrix} & \text{layer 1: matrix of shape (1, 2), output shape (1, 2)} \\
\mathbf{W}_2 &= \begin{bmatrix} w_3 \\ w_4 \end{bmatrix}& \text{layer 2: matrix of shape (2, 1), output shape (1, 1)} \\
\end{aligned}
$$

Assuming ReLUs and no biases, the entire equation for this network is
$y = \mathbf{W}_2 \cdot \text{ReLU}(\mathbf{W}_1 x) + b$ (final layer should always have a bias), that is:

$$
\begin{aligned}
\mathbf{W}_1 x & = \begin{bmatrix} w_1 \cdot x & \;\; w_2 \cdot x \end{bmatrix} \\
             y & = \begin{bmatrix}  w_3 \cdot \text{ReLU}(w_1 \cdot x) + w_4 \cdot \text{ReLU}(w_2 \cdot x) + b \end{bmatrix} \\
\end{aligned}
$$

This in fact reduces to:

$$
y = \begin{cases}
w_1 w_3 x + w_2 w_4 x + b  & \text{ if }  sgn(x) =   sgn(w_1) \text{ and } sgn(x)   = sgn(w_2) \\
w_1 w_3 x + b              & \text{ if }  sgn(x) =   sgn(w_1) \text{ and } sgn(x) \ne sgn(w_2) \\
w_2 w_4 x + b              & \text{ if }  sgn(x) \ne sgn(w_1) \text{ and } sgn(x)   = sgn(w_2) \\
b                          & x = 0 \\
\end{cases}
$$

In particular, without loss of generality we can take $w_1 < 0$ and $w_2 > 0$ and reduce this to:

$$
y = \begin{cases}
w_2 w_4 x + b  & \text{ if }  x \ge 0 \\
w_1 w_3 x + b  & \text{ if }  x  < 0 \\
\end{cases}
$$

# References
